---
title: "Final Project"
subtitle: "Project Proposal"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
self-contained: true
number-sections: true
---

## Project Proposal - Data {.unnumbered}

### Reading in Data {.unnumbered}

```{r}
library(tidyverse)
library(here)

child_mortality <- read.csv("child_mortality_0_5_year_olds_dying_per_1000_born.csv")
GDP_per_Cap <- read.csv("gdp_pcap.csv")
```

### Data at a Glance {.unnumbered}

```{r}
head(child_mortality)
head(GDP_per_Cap)


# mortality NAs 
sum(is.na(child_mortality))

# GDP NAs
sum(is.na(GDP_per_Cap))
```

## Let's Start Cleaning

### Limit Range
-   Some countries did not exist until 1950 (assumed they get their independence from a colonial power this year, so refine data frames to remove cols X1800 - X 1949

```{r}
# Refine child_mortality dataset
child_mortality <- child_mortality |>
  select(country, `X1950`:`X2100`)

# Refine GDP_per_Capita dataset
GDP_per_Cap <- GDP_per_Cap |>
  select(country, `X1950`:`X2100`)

#Renaming cols
child_mortality <- child_mortality|>
  rename_with(~ str_replace(., "^X", ""), .cols = everything())

GDP_per_Cap <- GDP_per_Cap |>
  rename_with(~ str_replace(., "^X", ""), .cols = everything())

```

### Pivoting Longer

```{r}
child_mortality_long <- child_mortality |>
  pivot_longer(
    cols = !country,
    names_to = "Year",
    values_to = "Mortality Rate"
  )

GDP_per_Cap_long <- GDP_per_Cap |>
  pivot_longer(
    cols = !country,
    names_to = "Year",
    values_to = "GDP Per Capita"
  )

```

### Convert to Numeric 
```{r}
# function converts numbers in "XXk" format to numeric
# eg "89k" -> 89,000
convert_k_to_number <- function(x) {
  if (grepl("k", x, fixed = TRUE)) {
    numeric_value <- as.numeric(sub("k", "", x)) * 1000
  } else {
    numeric_value <- as.numeric(x)
  }
  return(numeric_value)
}
```

```{r}
# converts mortality rate to numeric
child_mortality_long <- child_mortality_long |>
  filter(!is.na(child_mortality_long$`Mortality Rate`)) |>
  mutate(`Mortality Rate` = as.numeric(`Mortality Rate`))

# converts gdp to numeric
GDP_per_Cap_long <- GDP_per_Cap_long |>
  mutate(`GDP Per Capita` = map(`GDP Per Capita`, convert_k_to_number))
```

### Joining Datasets
```{r}
joined_df <- inner_join(child_mortality_long, GDP_per_Cap_long, join_by(country, Year))

joined_df
```

## Data Description

### Data & Variable Description

### Hypothesized Relationship

### Cleaning Process & Decisions
We decided to limit our examination to the 1950s-present because some countries
did not exist then, and therefore that took care of a lot of the NA values. We also
renamed the years to make it more easily readable by removing the X in front of the
number. We pivoted the data to be longer so it can be easily analyzed.
Then, we converted the values to be numeric as they represent quantitative data.
For the GDP, we had to convert some values that had a "k" in it to the actual multiple
of 1000.
