---
title: "Final Project"
subtitle: "Project Proposal"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
code-fold: true
self-contained: true
number-sections: true
---

## Project Proposal - Data {.unnumbered}

### Reading in Data {.unnumbered}

```{r}
library(tidyverse)
library(here)
child_mortality <- read_csv("child_mortality_0_5_year_olds_dying_per_1000_born.csv")
GDP_per_Cap <- read_csv("gdp_pcap.csv")
```
## Let's Start Cleaning
We decided to limit our examination to the 1950s-2020 because some countries
did not exist then, and therefore that took care of a lot of the NA values. The dataset also has years > 2020, but we wanted to look at how the past data points changed over time without the predicted values. We pivoted the data to be longer so it can be easily analyzed (on mortality rate and GDP per capita for each year).
Then, we converted the values to be numeric as they represent quantitative data.
For the GDP, we had to convert some values that had a "k" in it to the actual multiple
of 1000.

### Limit Range

```{r}
# Refine child_mortality dataset
child_mortality <- child_mortality |>
  select(country, `1950`:`2020`)

# Refine GDP_per_Capita dataset
GDP_per_Cap <- GDP_per_Cap |>
  select(country, `1950`:`2020`)
```


### Pivoting Longer

```{r}
child_mortality_long <- child_mortality |>
  pivot_longer(
    cols = !country,
    names_to = "Year",
    values_to = "Mortality Rate"
  )

GDP_per_Cap_long <- GDP_per_Cap |>
  pivot_longer(
    cols = !country,
    names_to = "Year",
    values_to = "GDP Per Capita"
  )

```

### Convert to Numeric 
```{r}
# function converts numbers in "XXk" format to numeric
# eg "89k" -> 89,000
convert_k_to_number <- function(x) {
  if (grepl("k", x, fixed = TRUE)) {
    numeric_value <- as.numeric(sub("k", "", x)) * 1000
  } else {
    numeric_value <- as.numeric(x)
  }
  return(numeric_value)
}
```

```{r}
# converts mortality rate to numeric
child_mortality_long <- child_mortality_long |>
  filter(!is.na(child_mortality_long$`Mortality Rate`)) |>
  mutate(`Mortality Rate` = as.numeric(`Mortality Rate`))

# converts gdp to numeric
GDP_per_Cap_long <- GDP_per_Cap_long |>
  mutate(`GDP Per Capita` = sapply(`GDP Per Capita`, convert_k_to_number))
```

### Joining Datasets
```{r}
joined_df <- inner_join(child_mortality_long, GDP_per_Cap_long, join_by(country, Year))
```

## Data Description
For our two data sets, we decided to look at the relationship between child mortality 
and GDP per capita. The data in the child mortality data set consists of a combination 
of multiple data sets; data from 1800 to 1950 was from v7 of Gapminder, data from 1950 
to 2016 was taken from UNIGME, a data collaboration project between UNICEF, WHO, 
UN Population Division, and the World Bank. Data from 1950 to 2100 was taken from UN POP, 
World Population Prospects 2019. The data in the GDP per capita data set is comprised of 
a combination of GDP per capita data from the World Bank, the Maddison Project Database, 
and the Penn World Table. We took these two data sets and joined them to create one data 
set that displays child mortality rate, GDP per capita, year, and country.

### Variable Description
The first variable in our joined data set is "country", which specifies the country that 
the following data describes. The next variable is "Year", which specifies the year in 
which the following data was collected. In our data, we limited our "Year" variable from 
1950-present because many countries did not exist and therefore had missing data prior to 
then. The "Mortality Rate" variable represents the number of deaths of children under 
5 years old per 1000 live births. Finally, the "GDP Per Capita" variable represents the 
gross domestic product per person. The data for "GDP Per Capita" was also adjusted for 
differences in purchasing power.

### Hypothesized Relationship
<<<<<<< HEAD
We think the an increase in GDP will correlate with a decrease in child mortality rates.Additionally we am expecting to see a dramatic decrease in child mortality over time. An article from the World Health Organization reported the rate has fallen over 59% since 1990. Additionally there will also be a massive increased in GDP due the industrialization and growth of industry. 
=======
We think that an increase in GDP will correlate with a decrease in child mortality rates as countries with higher GDP tend to have better resources
that can improve health like better living conditions, healthcare, etc.
Additionally we are expecting to see a dramatic decrease in child mortality over time. An [article](https://www.who.int/data/gho/data/themes/topics/topic-details/GHO/child-mortality-and-causes-of-death#:~:text=Since%201990%2C%20the%20global%20under,to%202.3%20million%20in%202021) from the World Health Organization reported the rate has fallen over 59% since 1990. 
>>>>>>> 1d1072100f0c5ebed2188d929b6f107748ae9f08

## Linear Regression

### 2.1 Data Visualization

Create two visualizations exploring the following:

The relationship between the two quantitative variables changing over time. Get creative!

The relationship between the two quantitative variables you are investigating.

In investigating this relationship, you may want to first average over the years for each country (i.e., one observational row per country since our independence condition would be violated).

#### Relationship between GDP and Child Mortality
```{r}

joined_df |>
  group_by(country) |>
  summarise(
    Avg_Mortality_Rate = mean(`Mortality Rate`, na.rm = TRUE),
    Avg_GDP_Per_Capita = mean(`GDP Per Capita`, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Avg_GDP_Per_Capita, y = Avg_Mortality_Rate)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Average GDP Per Capita", y = "Average Mortality Rate",
       title = "Average Mortality Rate vs Average GDP Per Capita by Country")

# not averaged for reference
ggplot(joined_df, aes(x = `GDP Per Capita`, y = `Mortality Rate`)) +
  geom_point(alpha = 0.2) +  # alpha for transparency if points overlap
  theme_minimal() +
  labs(x = "GDP Per Capita", y = "Mortality Rate",
       title = "Relationship Between Child Mortality and GDP Per Capita")

# lol yikes we are definitely keeping averages here

```


#### Relationship between GDP and Child Mortality Over Time
```{r}
joined_df |>
  group_by(Year) |>
  summarise(
    Avg_Mortality_Rate = mean(`Mortality Rate`, na.rm = TRUE),
    Avg_GDP_Per_Capita = mean(`GDP Per Capita`, na.rm = TRUE)
  ) |>
  ggplot() +
  geom_line(aes(x = as.numeric(Year), y = Avg_Mortality_Rate, color = "Mortality Rate")) +
  geom_line(aes(x = as.numeric(Year), y = Avg_GDP_Per_Capita, color = "GDP Per Capita")) +
  scale_color_manual(values = c("Mortality Rate" = "red", "GDP Per Capita" = "blue")) +
  theme_minimal() +
  labs(color = "Variable",
       title = "Global Average of Child Mortality Rate and GDP Per Capita Over Time",
       y = "Global Average (Rate or GDP)")

```

- ok, so mortality rate is showing at zero, even though this isn't the case. suspecting it's a scale issue.
- can try visualizations again but rescaled, or do log trans



### 2.2 Linear Regression
- Fit a simple linear regression model for the relationship between the variables you selected 
- Similar to your visualization, you will need to define which variable is the response (y) and which variable is the explanatory (x).
- we want to predict child mortality(y) based off of GDP (x)

### 2.3 Model Fit

Make a nicely formatted table that presents the following:

The variance in the response values
The variance in the fitted values from your regression model
The variance in the residuals from your regression model
Discuss the proportion of the variability in the response values that was accounted for by your regression model. What does this suggest about the “quality” of your model?
